Clinical advancement forecasting \thanks{\textit{\underline{Citation}}: \textbf{TBD}}

### 2024-04-12 paper/paper.tex
More figure captions

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{fancyhdr}       % header
\usepackage{graphicx}       % graphics
\usepackage[table, x11names, svgnames]{xcolor}
\usepackage{subfig}
\usepackage{float}
\usepackage{caption}
\usepackage[section]{placeins}
\usepackage{setspace}
\usepackage{overpic}

@@

%% Authors
\author{
  Authors TBD \\
  Related Sciences \\
%   %% examples of more authors
%    \And
%   Author3 \\
%   Affiliation \\
%   Univ \\
%   City\\
%   \texttt{email@email} \\
}

%% Body
\setlength{\parindent}{20pt}

@@

It has been well established that drugs with human genetic evidence linking their respective targets to indications in clinical trials are more likely to succeed \cite{Nelson2015-eg,King2019-rc,Minikel2023.06.23.23291765,Razuvayevskaya2023.02.07.23285407,PMID:30652614,PMID:24833294,PMID:35804044,PMID:37803084,PMID:36963162}, and to a lesser extent, that the same may also be true for single-cell transcriptomic evidence \cite{Dann2024.04.04.24305313}. This information has been used to devise target and target-disease ranking algorithms based primarily on a synthesis of multiple genetic signals alone \cite{PMID:38172303,Koscielny2017-rr,PMID:31253980}. It is also possible to expand the breadth of this genetic support to more targets and diseases based on knowledge graphs, protein interactions and/or disease ontologies \cite{PMID:33262371,Bao2022-bq,Sadler2023-xd,PMID:36087372,PMID:36823319}. To our knowledge, all such expansion methods identify a larger space of opportunities at the expense of expected success rates. This is not a focus of this work as we aim, instead, to establish a framework for identifying target-disease (TD) pairs with the very highest possible likelihood of success. We accomplish this by integrating human clinical, genetic/genomic, transcriptomic and proteomic data as well as cell/animal model evidence, pathway information and basic literature metrics from Open Targets \cite{Koscielny2017-rr} in statistical models that are trained on past clinical outcomes, evaluated primarily on their performance at upper extremes of predictions and then deployed on undeveloped TD pairs to establish an evidence-based ranking pipeline. This method can be contrasted with far more integrative methods that rely on neural and/or graph models over extensive knowledge graphs \cite{Paliwal2020-hr,PMID:33741907,pittala2020relationweighted,PMID:32750869}, which are more complex and difficult to interpret. We believe a desirable middle ground between these approaches and those that aim to combine many orthogonal indicators of success through expert knowledge in heuristic systems \cite{PMID:38404138,Koscielny2017-rr} would: 1) permit inclusion of many types of evidence, spanning hundreds or even thousands of putative orthogonal, causal predictors, 2) be highly interpretable, 3) support expert judgement where necessary, and 4) not require manual ranking/weighting schemes. This is the basis for our internal ranking platform, and such operational advantages have made it possible to reach or exceed the inclusivity of highly integrative methods without sacrificing interpretability, modularity and ease of deployment.

A substantial challenge inherent to building such a system is the need to account for the longitudinal nature of knowledge discovery in biomedicine. This is vital because any method that optimizes for likely \textit{future} clinical success based on \textit{historical} clinical outcomes may easily be biased by the non-random nature with which evidence can be absent otherwise. We use longitudinal evidence, i.e. evidence for which timing of its emergence can be determined, where appropriate when training and evaluating our methods before ultimately applying them to present-day evidence that is not temporalized. We discuss motivations, prior research and our own analysis on how important this problem is for each type of evidence in Section~\ref{sec:results_inflation}.

@@

\begin{figure}[!htb]
  \centering
  \captionsetup{width=.9\linewidth}
  \begin{overpic}[width=1\textwidth]{opportunity_funnel.pdf}
    \put(100, 53) {(a)}
    \put(100, 41) {(b)}
    \put(100, 19) {(c)}
    \put(50, 51){\color{black}\vector(0, -1){5}}
    \put(74, 39){\color{black}\vector(0, -1){4.5}}
    \put(13.5, 58){\sffamily \fontsize{8pt}{12pt}\selectfont Of 2.53M TD pairs with evidence linking them, 2.46M have never been in clinical trials.}
    \put(13.5, 35.8){\sffamily \fontsize{8pt}{12pt}\selectfont Of 2.46M undeveloped pairs, the top 0.8\% are 2.07x+ more likely to advance.}
    \put(13.5, 1.8){\sffamily \fontsize{8pt}{12pt}\selectfont Of 8,560 pairs 2.93x more likely to advance, 2,408 have clinical small molecule targets (e.g.).}
  \end{overpic}
  \caption{
    \textbf{Present-day target-disease pair opportunity funnel}.
    ... 
  }
  \label{fig:opportunity_funnel}
\end{figure}

\begin{figure}[!htb]
  \centering
  \captionsetup{width=.9\linewidth}
  \includegraphics[width=1\textwidth]{opportunity_summary.png}
  \caption{
    \textbf{Present-day target-disease pair counts by stage, likelihood of advancement and tractability}. The \textbf{stage} panel contains counts by maximum trial phase reached, the \textbf{threshold} panel contains counts of pairs with a RDG model score exceeding that of the associated benchmark for only undeveloped pairs, and the \textbf{tractability} panel shows pair frequencies among undeveloped pairs exceeding the EVA threshold that also have a HIGH tractability rating as defined in Supplementary~Table~\ref{tab:tractability_buckets}. This corresponds to targets that have all been in clinical development already, except for the \textbf{OC} modality in which case it indicates that a target has been approved. 
  }
  \label{fig:opportunity_summary}
\end{figure}

Figure~\ref{fig:opportunity_summary} shows the distribution of TD pair counts for select buckets across therapeutic areas as well as across the current maximum phase reached for any one pair. We find that there are $\sim$2,400 small-molecule-enabled, $\sim$1,400 antibody-enabled, and 14 PROTAC-enabled TD pairs with a probability of advancement that is nearly 3x other TD pairs based on the EVA threshold RR=2.93 in Figure~\ref{fig:relative_risk_by_limit}. Top antibody-enabled pairs are shown in Supplementary~Figure~\ref{fig:top_opportunity_predictions} along with their corresponding genetic and clinical support.

We also note that these TD pairs are identified using predictors that are not constrained by temporalization. In other words, the predictions used to identify \textbf{present-day} TD pairs with a high likelihood of clinical success are based on more evidence than what is available in other sections of this study. This is primarily because some evidence in Open Targets is not associated with publications. Ignoring that limitation is appropriate in this context.

@@

\begin{enumerate}
\item Mendelian randomization research is biased towards cardiovascular diseases as they have a disproportionate number of known, modifiable exposures \cite{PMID:36736292}
\item Putative protein interactions that do not result from genome-scale or otherwise unbiased assays result in an overrepresentation of successful drug targets in resources like STRING \cite{PMID:36370105}, thereby inflating the success of network expansion methods over these databases to identify such targets \cite{Sadler2023-xd}.
\item Transcript expression studies run in late-stage clinical trials for a single indication, e.g. \cite{PMID:27723281} linking SLE to IFN genes, are a degenerate indicator of advancement beyond earlier stage trials when the timing of this evidence is not accounted for.
\item Targets tested against more indications in clinical trials enrich for failures because the marginal cost of testing more indications decreases, but the evidence for these indications is often weaker \cite{PMID:33262371}.
\item Herding effects in pharma R\&D pipelines around particular drug targets are becoming increasingly clear over time \cite{PMID:37117303} and generate an excess of clinical evidence for those targets.
\end{enumerate}

We also note that the skew in basic drug target research towards those that already have rich annotations and well characterized molecular function \cite{PMID:29358745} as well as the disproportionate representation of particular target families in pharma R\&D pipelines \cite{PMID:27910877,PPR:PPR7029} and the fact that literature is well known to be biased away from negative results in general \cite{PMID:32893970} are all problematic.

@@

\begin{itemize}
\item \(A\) is the event that evidence for a TD pair arises after its first early-stage (phase 1 or 2) trial rather than before
\item \(B\) is the event that a TD pair advances into late-stage trials (phase 3 or 4)
\end{itemize}

@@

\begin{figure}[!htb]
  \centering
  \captionsetup{width=.9\linewidth}
  \includegraphics[width=1\textwidth]{evidence_inflation.pdf}
  \caption{
    \textbf{Clinical success drives evidence discovery}.  
  }
  \label{fig:evidence_inflation}
\end{figure}

@@

An important question, that remains difficult to answer, is how much target prioritization methods like the one explored in this study can improve the efficiency of drug discovery. One 2018 estimate suggests that improving success rates from phase 1 to approval by 1.4x (from 9.6\% to 13.8\%) would reduce the median R\&D cost for a \textbf{single drug} by \$480 million USD \cite{PMID:32125404}. If a 2x increase in success rates from phase 1 to approval could be expected when pursuing only genetically validated TD pairs \cite{Nelson2015-eg}, then this would increase to a \$686 million cost savings per drug. Pushing that even further to 3x or 4x increase by incorporating more lines evidence, as we have done here, would suggest a per-drug cost savings of \$1.03 billion and \$1.37 billion, respectively. These savings would only be realized for drugs tested against targets and diseases with sufficient supporting evidence, however, we show in Section~\ref{sec:results_opportunities} there are more than 8,000 such opportunities (i.e. untested, present-day TD pairs) where at least a 3x improvement in success rates could be expected.

% TODO: complete this section?
Another way to quantify the impact that employing superior target prioritization strategies could have is to consider what would happen if this was done across the entire pharmaceutical development industry. Our internal research shows that \$130B (92\%) of global pharma R\&D costs (\$150B) are expended on failed assets each year. This assumes an 8\% success rate that if improved by 3x to 24\%, would lead to a total cost savings of approximately \$22B annually ... (WIP)

@@

\begin{quotation}
  It is important to bear in mind therefore that what we are measuring when looking at historical trial outcomes is not an unbiased measure of any given gene's true disease associations, but rather a view on how useful a given evidence source or analytical method has been for choosing drug targets based on current and historical drug discovery practices. Dramatic changes in these practices in the future could render some of our conclusions obsolete, though the fundamental observation that genetic association itself is retained in molecular networks will remain valid.
\end{quotation}

@@

The target-specific features we derive come from the "target prioritisation" dataset \cite{OTtargetPrioritisation}. The only exception to this is a feature defining the maximum trial phase reached, which comes from the primary target-disease evidence dataset. Most scores in this dataset were not used because they cannot be temporalized. We make exceptions for those arising from large-scale, genome-wide assays like LOEUF from gnomAD \cite{PMID:32461654} and tissue expression scores from Human Protein Atlas \cite{PMID:25613900}. We retain a score derived from phenotype severity observed in mice knock-outs \cite{PMID:33231642} as well. All three of these feature sources are unique in that they are imputed differently from all other features when absent in model training. The "closed-world" assumption discussed in Section~\ref{sec:results_inflation} implies that a constant, zero imputation for most missing features is appropriate. This is not true for non-clinical, target-specific fields -- they are mean-imputed instead.

@@

\begin{enumerate}
  \item Open Targets does not currently provide drug approvals with associated dates. This is a part of the reason why we have chosen to focus only on advancement beyond phase 2, and not phase 3 as well. 
  \item Not all Open Targets evidence can be associated with publication dates. This evidence is ignored when used to construct longitudinal predictors; however, this limitation does not apply to Section~\ref{sec:results_opportunities}.
  \item The Open Targets composite score we use (OTS) is not equivalent to the overall composite score for TD pairs provided by Open Targets \cite{OTscoring}. The score they provide is not longitudinal. We create an approximation of it over time based on the same weighting scheme across evidence types, but without a harmonic sum aggregation function. Maximum scores for the same target, disease, year and evidence type combination are used instead.
  \item Estimating an absolute number of TD pairs associated with any criteria is difficult. In Section~\ref{sec:results_opportunities}, we present absolute counts of undeveloped TD pairs with tractable targets and a high likelihood of clinical success. These counts may be inflated by the fact that the associated targets and diseases are so sufficiently similar that they ought not to be counted as distinct examples. Defining a threshold for similarity to combat this problem is difficult and application-specific though, so we do not offer any solutions here.
  \item The source of OMIM associations we use is derived from a subset of associations available in EVA. This subset is defined by associations with a linked publication. This is only an approximation, but our internal research shows that 98\% of EVA records with linked publications also have at least one OMIM submission.
\end{enumerate}

@@

\begin{figure}[H]
  \centering
  \captionsetup{width=.9\linewidth}
  \includegraphics[width=1\textwidth]{dataset_statistics.png}
  \caption{
    \textbf{Training and evaluation dataset summary statistics}.\\\hspace{\textwidth}
    \textbf{balance}: percentage of TD pairs that eventually advanced beyond phase 2 \\\hspace{\textwidth} 
    \textbf{min\_year}: earliest year in which a TD pair has first advanced to phase 2 \\\hspace{\textwidth} 
    \textbf{n\_targets}: number of unique targets \\\hspace{\textwidth} 
    \textbf{n\_diseases}: number of unique diseases \\\hspace{\textwidth} 
    \textbf{n\_pairs}: number of TD pairs (equal to number of records) \\\hspace{\textwidth} 
    \textbf{n\_pairs\_wev}: number of TD pairs with directly associated evidence (i.e. not just target or disease evidence).
  }
  \label{fig:dataset_statistics}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \captionsetup{width=.9\linewidth}
  \includegraphics[width=1\textwidth]{relative_risk_core_features.pdf}
  \caption{
    \textbf{Performance of individual features and predictive scores as measured by relative risk}.
    RDG model results denoted by \colorbox{Gainsboro}{rdg@N} indicate performance for the \colorbox{Gainsboro}{N} top TD pairs. The same convention is used for \colorbox{Gainsboro}{literature} evidence and the \colorbox{Gainsboro}{time\_\_transition\_\_eq\_X} convention denotes RR estimates when the time since the phase 2 transition is equal to \colorbox{Gainsboro}{X} years. The \colorbox{Gainsboro}{omim}, \colorbox{Gainsboro}{eva}, and \colorbox{Gainsboro}{ot\_genetics\_portal} features correspond to the OMIM, EVA and OTG baselines of Figure~\ref{fig:relative_risk_by_limit}, respectively. All other features are assessed based on their existence. The counts along the origin indicate how many TD pairs were used to compute the RR numerator.
  }
  \label{fig:relative_risk_core_features}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \captionsetup{width=.9\linewidth}
  \includegraphics[width=1\textwidth]{relative_risk_static_features.pdf}
  \caption{
    \textbf{Relative risk scores for target/disease features}.
    The features ending with \colorbox{Gainsboro}{qtl\_Q} denote binary indicators constructed from cases where the feature meets or exceeds quantile \colorbox{Gainsboro}{Q} of its distribution. The features ending with \colorbox{Gainsboro}{gte\_X} denote indicators for when the feature meets or exceeds a specific value \colorbox{Gainsboro}{X}.
  }
  \label{fig:relative_risk_static_features}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \captionsetup{width=.9\linewidth}
  \includegraphics[width=1\textwidth]{relative_risk_by_ta.png}
  \caption{
    \textbf{Relative risk scores by method, benchmark and therapeutic area}.
    The \textbf{average} therapeutic area indicates mean values across all others except for \textbf{all}, which is an ungrouped estimate across all diseases regardless of therapeutic area.
  }
  \label{fig:relative_risk_by_ta}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \captionsetup{width=.9\linewidth}
  \includegraphics[width=1\textwidth]{relative_risk_dist_across_ta.pdf}
  \caption{
    \textbf{Relative risk distributions across select therapeutic areas}.
    P-values are computed from a one-sided Wilcoxon signed-rank test with the alternative that the RDG model RR averages across therapeutic areas exceed OTS averages.
  }
  \label{fig:relative_risk_dist_across_ta}
\end{figure}

>>>>

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{fancyhdr}       % header
\usepackage{graphicx}       % graphics
\usepackage{enumitem}       % lists
\usepackage[table, x11names, svgnames]{xcolor}
\usepackage{subfig}
\usepackage{float}
\usepackage{caption}
\usepackage[section]{placeins}
\usepackage{setspace}
\usepackage{overpic}

@@

%% Authors
\author{
  Eric Czech\thanks{Correspondence to Eric Czech (eric@related.vc).} \hspace{0.2cm} 
  Daniel Himmelstein \hspace{0.2cm} 
  Rafal Wojdyla \hspace{0.2cm} 
  \\ \\
  \textbf{Jack Milwid} \hspace{0.2cm} 
  \textbf{Adam Kolom} \hspace{0.2cm} 
  \textbf{Jeff Hammerbacher}
  \\ \\
  Related Sciences LLC
}

%% Body
\setlength{\parindent}{20pt} % Set paragraph indentation
\setlength{\fboxsep}{1pt} % Set top and bottom padding around a \colorbox

@@

It has been well established that drugs with human genetic evidence linking their respective targets to indications in clinical trials are more likely to succeed \cite{PMID:24833294,PMID:35804044,Razuvayevskaya2023.02.07.23285407,PMID:37803084,PMID:36963162}, typically by a factor of at least 2x \cite{Minikel2023.06.23.23291765,Nelson2015-eg,King2019-rc,PMID:30652614,Sadler2023-xd}. This information has been used to devise target and target-disease ranking algorithms based primarily on a synthesis of multiple genetic signals alone \cite{PMID:38172303,Koscielny2017-rr,PMID:31253980}. It is also possible to expand the breadth of this genetic support to more targets and diseases based on knowledge graphs, protein interactions and/or disease ontologies \cite{PMID:33262371,Bao2022-bq,Sadler2023-xd,PMID:36087372,PMID:36823319}. In contrast, we aim to establish a framework for identifying targets-disease (TD) pairs with the very highest possible likelihood of success in order to establish an evidence-based method to optimize research and development pipelines. This is accomplished in this study by integrating a subset of our evidence base from Open Targets \cite{Koscielny2017-rr} spanning human clinical, genomic, transcriptomic and proteomic data as well as cell and animal model evidence, pathway information, and basic literature metrics, in a simple statistical modeling framework.

This method contrasts with far more integrative methods that rely on neural and/or graph models over extensive knowledge graphs \cite{Paliwal2020-hr,PMID:33741907,pittala2020relationweighted,PMID:32750869}, which are more complex and often difficult to interpret. We believe a desirable middle ground between these two approaches and those that aim to combine many orthogonal indicators of success through expert knowledge in heuristic systems \cite{PMID:38404138,Koscielny2017-rr} would: 1) permit inclusion of many types of evidence from many sources to maximize inclusion of potentially causal, orthogonal predictive properties of targets and diseases, 2) be highly interpretable, 3) support expert judgement where necessary, and 4) not require manual ranking/weighting schemes.

A substantial challenge inherent to building such a system is the need to account for the longitudinal nature of knowledge discovery in biomedicine. This is vital because any method that optimizes for likely \textit{future} clinical success based on \textit{historical} clinical outcomes may easily be biased by the non-random nature with which evidence can be absent. By utilizing "temporalized" evidence -- i.e. evidence for which timing of its emergence can be determined -- and outcomes when training our methods before ultimately applying them to present-day evidence with no restrictions on timing, we reduce the risk of information leaks and inflation of performance estimates. We discuss motivations, prior research and our own analysis on how important this problem is for each source in Section~\ref{sec:results_inflation}.

@@

\begin{figure}[!htb]
  \centering
  \captionsetup{width=.9\linewidth}
  \begin{overpic}[width=1\textwidth]{opportunity_funnel.pdf}
    \put(100, 53) {(a)}
    \put(100, 41) {(b)}
    \put(100, 19) {(c)}
    \put(50, 51){\color{black}\vector(0, -1){5}}
    \put(74, 39){\color{black}\vector(0, -1){4.5}}
    \put(13.5, 58){\sffamily \fontsize{8pt}{12pt}\selectfont Of 2.53M TD pairs with evidence linking them, 2.46M have never been in clinical trials.}
    \put(13.5, 35.8){\sffamily \fontsize{8pt}{12pt}\selectfont Of 2.46M undeveloped pairs, the top 0.8\% are 2.07x+ more likely to advance.}
    \put(13.5, 1.8){\sffamily \fontsize{8pt}{12pt}\selectfont Of 8,560 pairs 2.93x more likely to advance, 2,408 have clinical small molecule targets (e.g.).}
  \end{overpic}
  \caption{
    \textbf{Present-day target-disease pair opportunity funnel}.
    (a) All 2.53M TD pairs in Open Targets with evidence of some kind grouped by the maximum clinical trial phase reached for each as of the present day. The \colorbox{Gainsboro}{Undeveloped} pairs have never been tested in clinical trials. TD pair counts are shown in parentheses. Labels for phase 1, 3 and 4 trials are omitted for brevity.
    (b) All 2.46M undeveloped TD pairs for which RDG model predictions exceed the thresholds determined by the corresponding genetic benchmarks in Figure~\ref{fig:relative_risk_by_limit}. The \colorbox{Gainsboro}{Medium} threshold was calibrated by genetic support across OTG (i.e. GWAS evidence). The associated RR of 2.07 implies that pairs with RDG predictions above this threshold are at least 2.07 more likely to advance clinically than those below the threshold. The same is true for the \colorbox{Gainsboro}{High} and \colorbox{Gainsboro}{Highest} thresholds, which correspond to thresholds determined by EVA and OMIM, respectively. These groupings are not mutually exclusive, e.g. the 20,821 TD pairs above the \colorbox{Gainsboro}{Medium} threshold are a superset of the pairs above the \colorbox{Gainsboro}{High} threshold.
    (c) All 8,560 undeveloped TD pairs with an expected clinical advancement rate of at least 2.93x grouped by tractability modality and confidence level. The \colorbox{Gainsboro}{Other} modality corresponds to any modality not shown that is associated with a target tested in clinical trials previously. The \colorbox{Gainsboro}{High} confidence grouping for this category means those targets have approved drugs. For each other modality, the \colorbox{Gainsboro}{High} confidence level corresponds to \textbf{any} level of clinical validation for drugs of that modality, not just approval. See Supplementary~Table~\ref{tab:tractability_buckets} for the definition of all modality confidence levels.
  }
  \label{fig:opportunity_funnel}
\end{figure}

Figure~\ref{fig:opportunity_funnel} demonstrates this process visually, starting from the space of all present-day TD pairs. This funnel implies there are thousands of TD pairs that 1) have never been tested in clinical trials before, 2) have a 3x (RR=2.93) higher likelihood of advancing past phase 2 trials and 3) have a target that \textbf{has} been tested clinically (in at least phase 1 trials) for one or more of the 
modalities highlighted. In other words, there are $\sim$2,400 small-molecule-enabled, $\sim$1,400 antibody-enabled, and 14 PROTAC-enabled TD pairs at the bottom of the funnel where the implied opportunity is greatest. Examples of antibody-enabled TD pairs are shown in Supplementary~Figure~\ref{fig:top_opportunity_predictions} along with their corresponding genetic and clinical support.

For similar statistics by therapeutic area, rather than across all of them as in Figure~\ref{fig:opportunity_funnel}, see Supplementary~Figure~\ref{fig:opportunity_summary}. This also shows TD pair counts for trial phases omitted from Figure~\ref{fig:opportunity_funnel}.

Lastly, we note that these TD pairs are identified using predictors that are not constrained by temporalization. In other words, the predictions used to identify \textbf{present-day} TD pairs with a high likelihood of clinical success are based on more evidence than what is available in other sections of this study. This is primarily because some evidence in Open Targets is not associated with publications. Avoiding that limitation is appropriate in this context.

@@

\begin{enumerate}[topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
\item Mendelian randomization research is biased towards cardiovascular diseases as they have a disproportionate number of known, modifiable exposures \cite{PMID:36736292}
\item Putative protein interactions that do not result from genome-scale or otherwise unbiased assays result in an overrepresentation of successful drug targets in resources like STRING \cite{PMID:36370105}, thereby inflating the success of network expansion methods over these databases to identify such targets \cite{Sadler2023-xd}.
\item Transcript expression studies run in late-stage clinical trials for a single indication, e.g. \cite{PMID:27723281} linking SLE to IFN genes, are a degenerate indicator of advancement beyond earlier stage trials when the timing of this evidence is not accounted for.
\item Targets tested against more indications in clinical trials enrich for failures because the marginal cost of testing more indications decreases, but the evidence for these indications is often weaker \cite{PMID:33262371}.
\item Herding effects in pharma R\&D pipelines around particular drug targets are becoming increasingly clear over time \cite{PMID:37117303} and generate an excess of clinical evidence for those targets.
\end{enumerate}

We also note that the skew in basic drug target research towards those that already have rich annotations and well characterized molecular function \cite{PMID:29358745}, the disproportionate representation of particular target families in pharma R\&D pipelines \cite{PMID:27910877,PPR:PPR7029} and the fact that literature is well known to be biased away from negative results in general \cite{PMID:32893970} are all problematic.

@@

\begin{itemize}[topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
\item \(A\) is the event that evidence for a TD pair arises after its first early-stage (phase 1 or 2) trial rather than before
\item \(B\) is the event that a TD pair advances into late-stage trials (phase 3 or 4)
\end{itemize}

@@

\begin{figure}[!htb]
  \centering
  \captionsetup{width=.9\linewidth}
  \begin{overpic}[width=1\textwidth]{evidence_inflation.pdf}
    \put(73, 42) {(a)}
    \put(73, 16) {(b)}
  \end{overpic}
  \caption{
    \textbf{Inflation risk by feature}.  
    (a) Feature "inflation risk", for more frequent features, reported as relative risk point estimates (dots) and Katz 99\% confidence intervals (bars). These features are "frequent" in the sense that they have enough TD pairs where evidence emerged before \textbf{and} after a phase 2 transition to power a relatively narrow confidence interval.
    (b) Rare feature inflation risk with the same definition as (a) and a wider range on the x-axis. All point estimates and confidence intervals are clamped to the range (0, 10) to minimize scale distortion from extreme estimates.
  }
  \label{fig:evidence_inflation}
\end{figure}

@@

An important question, that remains difficult to answer, is how much target prioritization methods like the one explored in this study can improve the efficiency of drug discovery. One 2018 estimate suggests that improving success rates from phase 1 to approval by 1.4x (from 9.6\% to 13.8\%) would reduce the median R\&D cost for a \textbf{single drug} by \$480 million USD \cite{PMID:32125404}. If a 2x increase in success rates from phase 1 to approval could be expected when pursuing only genetically validated TD pairs \cite{Nelson2015-eg}, then this would increase to a \$686 million cost savings per drug. Pushing that even further to 3x or 4x increase by incorporating more lines evidence, as we have done here, would suggest a per-drug cost savings of \$1.03 billion and \$1.37 billion, respectively. These savings would only be realized for drugs tested against targets and diseases with sufficient supporting evidence; however, we show in Section~\ref{sec:results_opportunities} there are more than 8,000 such opportunities (i.e. untested, present-day TD pairs) where at least a 3x improvement in success rates could be expected.

Another way to quantify the impact that employing superior target prioritization strategies could have, is to consider what would happen if this was done across the entire pharmaceutical development industry. Our internal research shows that \$130B (92\%) of global pharma R\&D costs (\$150B) are expended on failed assets each year. This assumes an 8\% success rate that if improved by 3x to 24\%, would lead to a total cost savings of approximately \$22B annually ... (TODO: finish with Adam)

@@

\begin{quote}
  It is important to bear in mind therefore that what we are measuring when looking at historical trial outcomes is not an unbiased measure of any given gene's true disease associations, but rather a view on how useful a given evidence source or analytical method has been for choosing drug targets based on current and historical drug discovery practices. Dramatic changes in these practices in the future could render some of our conclusions obsolete, though the fundamental observation that genetic association itself is retained in molecular networks will remain valid.
\end{quote}

@@

The target-specific features we derive come from the "target prioritisation" dataset \cite{OTtargetPrioritisation} in Open Targets. The only exception to this is a feature defining the maximum trial phase reached, which comes from the primary target-disease evidence dataset. Most scores in this dataset were not used because they cannot be temporalized. We make exceptions for those arising from large-scale, genome-wide assays like LOEUF from gnomAD \cite{PMID:32461654} and tissue expression scores from Human Protein Atlas \cite{PMID:25613900}. We retain a score derived from phenotype severity observed in mice knock-outs \cite{PMID:33231642} as well. All three of these feature sources are unique in that they are imputed differently from all other features when absent in model training. The "closed-world" assumption discussed in Section~\ref{sec:results_inflation} implies that a constant, zero imputation for most missing features is appropriate. This is not true for non-clinical, target-specific fields -- they are mean-imputed instead.

@@

\begin{enumerate}[topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  \item Open Targets does not currently provide drug approvals with associated dates. This is a part of the reason why we have chosen to focus only on advancement beyond phase 2, and not phase 3 as well. 
  \item Not all Open Targets evidence can be associated with publication dates. This evidence is ignored when used to construct longitudinal predictors; however, this limitation does not apply to Section~\ref{sec:results_opportunities}.
  \item The Open Targets composite score we use (OTS) is not equivalent to the overall composite score for TD pairs provided by Open Targets \cite{OTscoring}. The score they provide is not longitudinal. We create an approximation of it over time based on the same weighting scheme across evidence types, but without a harmonic sum aggregation function. Maximum scores for the same target, disease, year and evidence type combination are used instead.
  \item Estimating an absolute number of TD pairs associated with any criteria is difficult. In Section~\ref{sec:results_opportunities}, we present absolute counts of undeveloped TD pairs with tractable targets and a high likelihood of clinical success. These counts may be inflated by the fact that the associated targets and diseases are so sufficiently similar that they ought not to be counted as distinct examples. Defining a threshold for similarity to combat this problem is difficult and application-specific though, so we do not offer any solutions here.
  \item The source of OMIM associations we use is derived from a subset of associations available in EVA. This subset is defined by associations with a linked publication. This is only an approximation, but our internal research shows that 98\% of EVA records with linked publications also have at least one OMIM submission.
\end{enumerate}

@@

\begin{figure}[H]
  \centering
  \captionsetup{width=.9\linewidth}
  \includegraphics[width=1\textwidth]{dataset_statistics.png}
  \caption{
    \textbf{Training and evaluation dataset summary statistics}.\\\hspace{\textwidth}
    \colorbox{Gainsboro}{balance}: percentage of TD pairs that eventually advanced beyond phase 2 \\\hspace{\textwidth} 
    \colorbox{Gainsboro}{min\_year}: earliest year in which a TD pair has first advanced to phase 2 \\\hspace{\textwidth} 
    \colorbox{Gainsboro}{n\_targets}: number of unique targets \\\hspace{\textwidth} 
    \colorbox{Gainsboro}{n\_diseases}: number of unique diseases \\\hspace{\textwidth} 
    \colorbox{Gainsboro}{n\_pairs}: number of TD pairs (equal to number of records) \\\hspace{\textwidth} 
    \colorbox{Gainsboro}{n\_pairs\_wev}: number of TD pairs with directly associated evidence (i.e. not just target or disease evidence).
  }
  \label{fig:dataset_statistics}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \captionsetup{width=.9\linewidth}
  \includegraphics[width=1\textwidth]{relative_risk_core_features.pdf}
  \caption{
    \textbf{Performance of individual features and predictive scores as measured by relative risk}.
    RR estimates are shown as dots with bars indicating 90\% Katz confidence intervals.
    RDG model results denoted by \colorbox{Gainsboro}{rdg@N} indicate performance for the \colorbox{Gainsboro}{N} top TD pairs. The same convention is used for \colorbox{Gainsboro}{literature} evidence and the \colorbox{Gainsboro}{time\_\_transition\_\_eq\_X} convention denotes RR estimates when the time since the phase 2 transition is equal to \colorbox{Gainsboro}{X} years. The \colorbox{Gainsboro}{omim}, \colorbox{Gainsboro}{eva}, and \colorbox{Gainsboro}{ot\_genetics\_portal} features correspond to the OMIM, EVA and OTG baselines of Figure~\ref{fig:relative_risk_by_limit}, respectively. All other features are assessed based on their existence. The counts along the origin indicate how many TD pairs were used to compute the RR numerator.
  }
  \label{fig:relative_risk_core_features}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \captionsetup{width=.9\linewidth}
  \includegraphics[width=1\textwidth]{relative_risk_static_features.pdf}
  \caption{
    \textbf{Relative risk scores for target-only and disease-only features}.
    RR estimates are shown as dots with bars indicating 90\% Katz confidence intervals.
    The features ending with \colorbox{Gainsboro}{qtl\_Q} denote binary indicators constructed from cases where the feature meets or exceeds quantile \colorbox{Gainsboro}{Q} of its distribution. The features ending with \colorbox{Gainsboro}{gte\_X} denote indicators for when the feature meets or exceeds a specific value \colorbox{Gainsboro}{X}. The counts along the origin indicate how many TD pairs were used to compute the RR numerator.
  }
  \label{fig:relative_risk_static_features}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \captionsetup{width=.9\linewidth}
  \includegraphics[width=1\textwidth]{relative_risk_by_ta.png}
  \caption{
    \textbf{Relative risk scores by method, benchmark and therapeutic area}.
    The \colorbox{Gainsboro}{average} therapeutic area denotes mean values across all other therapeutic areas except for \colorbox{Gainsboro}{all}, which is an ungrouped estimate across all diseases regardless of therapeutic area. The \colorbox{Gainsboro}{n\_pairs} field indicates the number of TD pairs used to compute the RR estimate in each cell. The \colorbox{Gainsboro}{OTS@N} and \colorbox{Gainsboro}{RDG@N} columns indicate RR estimates for the top \colorbox{Gainsboro}{N} TD pairs from the OTS and RDG methods, respectively.
  }
  \label{fig:relative_risk_by_ta}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \captionsetup{width=.9\linewidth}
  \includegraphics[width=1\textwidth]{relative_risk_dist_across_ta.pdf}
  \caption{
    \textbf{Relative risk distributions across select therapeutic areas}.
    Relative risk point estimates for each of 13 therapeutic areas are shown as dots with boxplots underneath summarizing their distribution. P-values on the bottom margin of the top facet are computed from a one-sided Wilcoxon signed-rank test with the alternative that the RDG model RR average across therapeutic areas exceeds the OTS average at a given number of top ranking TD pairs (on the x-axis).
  }
  \label{fig:relative_risk_dist_across_ta}
\end{figure}
